Programming Language Foundations in Agda

### 2019-07-16 papers/scp/PLFA.tex
minor edits

However, after five years of experience, Phil came to the conclusion
that Coq is not the best vehicle.  Too much of the course needs to
focus on learning tactics for proof derivation, to the cost of
learning the fundamentals of programming language theory.  Every
concept has to be learned twice: e.g., both the product data type, and
the corresponding tactics for introduction and elimination of
conjunctions.  The rules Coq applies to generate induction hypotheses
can sometimes seem mysterious.  While the \texttt{notation} construct
permits pleasingly flexible syntax, it can be confusing that the same
concept must always be given two names, e.g., both
\texttt{subst~N~x~M} and \texttt{N~[x~:=~M]}.  Names of tactics are
sometimes short and sometimes long; naming conventions in the standard
library can be wildly inconsistent.  \emph{Propositions as types} as a
foundation of proof is present but hidden.

@@

PLFA is aimed at students in the last year of an undergraduate
honours programme or the first year of a master or doctorate degree.
It aims to teach the fundamentals of operational semantics of
programming languages, with simply-typed lambda calculus as the
central example.  The textbook is written as a literate script in Agda.
As with SF, the hope is that using
a proof assistant will make the development more concrete
and accessible to students, and give them rapid feedback to find
and correct misapprehensions.

The book is broken into three parts. The first part, Logical Foundations,
develops the needed formalisms.  The second part, Programming Language
Foundations, introduces basic methods of operational semantics.
The third part, Denotational Semantics, introduces a simple
model of semantics and its properties.
(SF is divided into books, the first two of which have the same names
as the first two parts of PLFA, and cover similar material.)
Part~I and Part~II up to Untyped were written by Phil,
Part~II from Substitution and Part~III were written by Jeremy.

@@

SF has a third volume, written by Andrew Appel, on Verified Functional
Algorithms. I'm not sufficiently familiar with that volume to have a view on
whether it would be easy or hard to cover that material in Agda. And SF recently
added a fourth volume on random testing of Coq specifications using QuickChick.
There is currently no tool equivalent to QuickChick available for Agda.

@@

PLFA puts the formal statements first, followed by informal explanation.
PLFA introduces an auxiliary relation \texttt{Progress} to capture
progress; an exercise (not shown) asks the reader to show it isomorphic
to the usual formulation with a disjunction and an existential.
Layout is used to present the auxiliary relation in inference rule form.
In Agda, any line beginning with two dashes is treated as a comment, making
it easy to use a line of dashes to separate hypotheses from conclusion
in inference rules.  The proof of proposition \texttt{progress} (the different
case making it a distinct name) is layed out carefully. The neat
indented structure emphasises the case analysis, and all right-hand
sides line-up in the same column.  Our hope as authors is that students
will read the formal proof first, and use it as a tabular guide
to the informal explanation that follows.

SF puts the informal explanation first, followed by the formal proof. The text
hides the formal proof script under an icon; the figure shows what appears when
the icon is expanded. As teachers, we were aware that students might skip the
formal proof on a first reading, and we would have to hope the students would
return to it and step through it with an interactive tool in order to make it
intelligible.  We expect the students skipped over many such proofs.  This
particular proof forms the basis for a question of the mock exam and the past
exams, so we expect most students will actually look at this one if not all the
others.

@@

In Coq, interaction consists of stepping through a proof script, at
each point examining the current goal and the variables currently in
scope, and executing a new command in the script.  Tactics are a whole
sublanguage, which must be learned in addition to the language for
expressing specifications.  There are many tactics one can invoke in
the script at each point; one menu in CoqIDE lists about one hundred
tactics one might invoke, some in alphabetic submenus.  A Coq
script presents the specification proved and the tactics executed.
Interaction is recorded in a script, which the students
may step through at their leisure.  SF contains some prose descriptions
of stepping through scripts, but mainly contains scripts that students
are encouraged to step through on their own.

In Agda, interaction consists of writing code with holes, at each
point examining the current goal and the variables in scope, and
typing code or executing an Emacs command.  The number of commands
available is much smaller than with Coq, the most important ones being
to show the type of the hole and the types of the variables in scope;
to check the code; to do a case analysis on a given variable; or to
guess how to fill in the hole with constructors or variables in scope.
An Agda proof consists of typed code.  The interaction is \emph{not}
recorded.  Students may recreate it by commenting out bits of code and
introducing a hole in their place.   PLFA contains some prose descriptions
of interactively building code, but mainly contains code that students
can read.  They may also introduce holes to interact with the code, but
we expect this will be rarer than with SF.

@@

Because the course is taught using a proof assistant, it is important
that students have access to a proof assistant during the exam.
Students are told in advance that they are expected to get perfect on
the exam, and that they will have to study hard to achieve this level.
Given that the goal of formal methods is to avoid error, we believe a
pedagogical purpose is served by telling the students that they are
expected to achieve perfection and making it possible for them to do
so.  Students are given two opportunities to practice in the run up to
the exam, a `mock' exam given in class under exam conditions (two
hours online), and before that a `mock mock' exam as coursework (in
their own time, encouraged to ask questions, tasked to do all three
questions rather than two of three).

>>>>

However, after five years of experience, Philip came to the conclusion
that Coq is not the best vehicle.  Too much of the course needs to
focus on learning tactics for proof derivation, to the cost of
learning the fundamentals of programming language theory.  Every
concept has to be learned twice: e.g., both the product data type, and
the corresponding tactics for introduction and elimination of
conjunctions.  The rules Coq applies to generate induction hypotheses
can sometimes seem mysterious.  While the \texttt{notation} construct
permits pleasingly flexible syntax, it can be confusing that the same
concept must always be given two names, e.g., both
\texttt{subst~N~x~M} and \texttt{N~[x~:=~M]}.  Names of tactics are
sometimes short and sometimes long; naming conventions in the standard
library can be wildly inconsistent.  \emph{Propositions as types} as a
foundation of proof is present but hidden.

@@

PLFA is aimed at students in the last year of an undergraduate
honours programme or the first year of a master or doctorate degree.
It aims to teach the fundamentals of semantics of
programming languages, with simply-typed and untyped lambda calculi as the
central examples.  The textbook is written as a literate script in Agda.
As with SF, the hope is that using
a proof assistant will make the development more concrete
and accessible to students, and give them rapid feedback to find
and correct misapprehensions.

The book is broken into three parts. The first part, Logical Foundations,
develops the needed formalisms.  The second part, Programming Language
Foundations, introduces basic methods of operational semantics.
The third part, Denotational Semantics, introduces a simple
model of the lambda calculus and its properties.
(SF is divided into books, the first two of which have the same names
as the first two parts of PLFA, and cover similar material.)
Part~I and Part~II up to Untyped were written by Phil,
Part~II from Substitution and Part~III were written by Jeremy.

@@

SF has a third volume, written by Andrew Appel, on Verified Functional
Algorithms. I'm not sufficiently familiar with that volume to have a view on
whether it would be easy or hard to cover that material in Agda. And SF recently
added a fourth volume on random testing of Coq specifications using QuickChick.
There is currently no tool equivalent to QuickChick for Agda.

@@

PLFA puts the formal statements first, followed by informal explanation.
PLFA introduces an auxiliary relation \texttt{Progress} to capture
progress; an exercise (not shown) asks the reader to show it isomorphic
to the usual formulation with a disjunction and an existential.
Layout is used to present the auxiliary relation in inference rule form.
In Agda, any line beginning with two dashes is treated as a comment, making
it easy to use a line of dashes to separate hypotheses from conclusion
in inference rules.  The proof of proposition \texttt{progress} (the different
case making it a distinct name) is layed out carefully. The neat
indented structure emphasises the case analysis, and all right-hand
sides line-up in the same column.  Our hope as authors is that students
read the formal proof first, and use it as a tabular guide
to the informal explanation that follows.

SF puts the informal explanation first, followed by the formal proof. The text
hides the formal proof script under an icon; the figure shows what appears when
the icon is expanded. As teachers, we were aware that students might skip the
formal proof on a first reading, and we have to hope the students 
return to it and step through it with an interactive tool in order to make it
intelligible.  We expect the students skipped over many such proofs.  This
particular proof forms the basis for a question of the mock exam and the past
exams, so we expect most students will look at this one if not all the
others.

@@

In Coq, interaction consists of stepping through a proof script, at
each point examining the current goal and the variables in
scope, and executing a new command in the script.  Tactics are a whole
sublanguage, which must be learned in addition to the language for
expressing specifications.  There are many tactics one can invoke in
the script at each point; one menu in CoqIDE lists about one hundred
tactics one might invoke, some in alphabetic submenus.  A Coq
script presents the specification proved and the tactics executed.
Interaction is recorded in a script, which the students
may step through at their leisure.  SF contains some prose descriptions
of stepping through scripts, but mainly contains scripts that students
are encouraged to step through on their own.

In Agda, interaction consists of writing code with holes, at each
point examining the current goal and the variables in scope, and
typing code or executing an Emacs command.  The number of commands
available is much smaller than with Coq, the most important ones being
to show the type of the hole and the types of the variables in scope;
to check the code; to do a case analysis on a given variable; or to
guess how to fill in the hole with constructors or variables in scope.
An Agda proof consists of typed code.  The interaction is \emph{not}
recorded.  Students may recreate it by commenting out bits of code and
introducing a hole in their place, but we expect this will be rarer
than with SF.  PLFA contains some prose descriptions of interactively
building code, but mainly contains code that students can read.

@@

Because the course is taught using a proof assistant, it is important
that students have access to a proof assistant during the exam.
Students are told in advance that they are expected to get perfect on
the exam, and that they will have to study hard to achieve it.
Given that the goal of formal methods is to avoid error, we believe a
pedagogical purpose is served by telling the students that they are
expected to achieve perfection and making it possible for them to do
so.  Students are given two opportunities to practice before
the exam, a `mock' exam given in class under exam conditions (two
hours online), and before that a `mock mock' exam as coursework (in
their own time, encouraged to ask questions, tasked to do all three
questions rather than two of three).
